---
title: "Assignment 3: Data Exploration"
author: "Taro Katayama, Section #1"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration. 

## Directions
1. Change "Student Name, Section #" on line 3 (above) with your name and section number.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "FirstLast_A03_DataExploration.Rmd") prior to submission.

The completed exercise is due on <>.


## Set up your R session

1. Check your working directory, load necessary packages (tidyverse), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. **Be sure to add the `stringsAsFactors = TRUE` parameter to the function when reading in the CSV files.**
```{r,  message = FALSE}
getwd()
setwd("/Users/tarokatayama/Desktop/Duke_Semester_2/Environmental_data_analytics/R_Projects/Environmental_Data_Analytics_2022/")
getwd()
library(tidyverse)
library(ggplot2)
Neonics<- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)
view(Neonics)
Litter<- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)
View(Litter)
```

## Learn about your system

2. The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: We might be interested in the ecotoxicology of neonicotinoids for a variety of reasons. 1) we might be interested in the effects of neonicotinoids on human health. 2) we might be interested in the effects of neonicotinoids on wildlife, specifically non-target animals. 3) we might be concerned with the effectiveness of neonicotinoids on target species.

3. The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. 

> Answer: We might be interested in studying litter and woody debris that falls to the ground because they may indicate which species of insect lives within the leaf litter. For example, the spiders may like a certain type of leaf litter, and are found most ofetn there. It will be easier to identify the ideal habitat for these insect species, and apply the insecticide appropriately.

4. How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here: 

> Answer: 
*Litter and woody debris was sampled at terrestrial NEON sites that contain woody vegetation that is greater than 2m tall.
*In sites with forested tower airsheds, the samples took place in 20 40mx40m plots.
*In sites with low-saturated vegetation over the tower airshed, the samples take place in 40mx40m tower plots + 26 20mx20m plots.



## Obtain basic summaries of your data (Neonics)

5. What are the dimensions of the dataset?

```{r}
dim(Neonics)
```
>Answer: 30 columns and 4623 rows

6. Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?
```{r}
summary(Neonics$Effect)

```

> Answer: population (1803 cases), followed by Mortality (1493), and Behavior (360) is the most common effects. These effects are of interest because they give you a sense of how effective the insecticide is to an assortment of insects. 

7. Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.

```{r}
summary(Neonics$Species.Common.Name)
```

> Answer: Honey Bee (667), Parasitic Wasp (285), Buff Tailed Bumblebee (183), Carniolan Honey Bee (152), Bumble Bee (140) were the most commonly studied speices in the dataset. They are all bee/wasp species. They might "bee" of interest because bees are very important species that are responsible for a number of ecological services such as pollination and honey production. When developing insecticides, they may want to create one that does not have adverse effects on the important bee species. 

8. Concentrations are always a numeric value. What is the class of Conc.1..Author. in the dataset, and why is it not numeric?

```{r}

class(Neonics$Conc.1..Author.)
help("factor")
class(Neonics$Test.Location)
```

> Answer: The class is a factor. It is not numeric because the class Conc.1..Author is categorical data that has a limited number of different values. Categorical data is most appropriately classed as a factor. 

## Explore your data graphically (Neonics)

9. Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year. 

```{r}
ggplot(Neonics) + 
  geom_freqpoly(aes(x=Publication.Year))
```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.
```{r}
ggplot(Neonics) + 
  geom_freqpoly(aes(x=Publication.Year, color=Test.Location), bin=50)
    
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test locations are 'labs', followed by 'field natural', followed by 'field artificial'. The general trend shows that tests increase over time, with lab test locations peaking around 2015. The field natural test location on the other hand peaks around 2009, and then drops off after that. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

```{r}
ggplot(Neonics, aes(x=Endpoint)) +
  geom_bar()
```

> Answer: The two most common endpoints are 'NOEL' and 'LOEL'. 'NOEL' stands for 'no observable effect level', which means the highest dose given without producing a significant effect compared to the control. 'LOEL' stands for 'Lowest observable effect level', which means lowest dose producing effects that were significantly different from control.  
>

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018. 
```{r}
class(Litter$collectDate)

Litter$collectDate<- as.Date(Litter$collectDate, format= "%Y-%m-%d")

class(Litter$collectDate)

View(Litter)

unique(Litter$collectDate)
```
> Answer: Using the 'unique' function, I found out that litter was sampled on 08/02/2018, and 08/30/2018

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
unique(Litter$plotID)
summary(Litter$plotID)
```

> Answer: There are 12 unique plots. Unique is different from summary because unique only gives you unique plots at Niwot Ridge. Summary gives you unique plots but also gives you the number of samples from each plot. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.
```{r}
ggplot(Litter, aes(x=functionalGroup))+
  geom_bar()
```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup. 
```{r}
ggplot(Litter) +
  geom_boxplot(aes(x=dryMass, y=functionalGroup))

ggplot(Litter) +
  geom_violin(aes(x=dryMass, y=functionalGroup))


```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: Boxplot is more effective because in this case, the median for the data is very close to zero for most functional groups, and is not visually represented well in violin plots.

What type(s) of litter tend to have the highest biomass at these sites? 

> Answer: Needles has the highest biomass, followed by mixed. 