---
title: "Assignment 5: Data Visualization"
author: "Taro Katayama"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Monday, February 14 at 7:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the [`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
getwd()
library(tidyverse)
library(cowplot)
NTL_LTER<- read.csv("../Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
NEON_LITTER<- read.csv("../Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv")

#2 
NTL_LTER$sampledate<- as.Date(NTL_LTER$sampledate, "%Y-%m-%d")
NEON_LITTER$collectDate<- as.Date(NEON_LITTER$collectDate, "%Y-%m-%d")
```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
Taro_Theme <- theme_classic(base_size = 10) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "bottom")
theme_set(Taro_Theme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and `ylim()`).

```{r}
#4
PeterPaulPlot1<-
  ggplot(NTL_LTER, aes(x=tp_ug, y=po4, color= lakename))+
  geom_point(size=0.5)+
  geom_smooth(method = lm, color='black')+
   xlim(0, 150) +
  ylim(0, 50)+
  xlab("Total Phosphorus")+
  ylab("Phosphate")
print(PeterPaulPlot1)
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
#5

NTL_LTER$month<- as.factor(NTL_LTER$month)

peterpaulplot2<-
  ggplot(NTL_LTER, aes(x=month, y=temperature_C))+
  geom_boxplot(aes(color= lakename))+
  ylab("Temperature C")
print(peterpaulplot2)

peterpaulplot3<-
  ggplot(NTL_LTER, aes(x=month, y=tp_ug))+
  geom_boxplot(aes(color=lakename))+
  ylab("TP")
  print(peterpaulplot3)

peterpaulplot4<-
  ggplot(NTL_LTER, aes(x=month, y=tn_ug))+
  geom_boxplot(aes(color=lakename))+
  ylab("TN")
print(peterpaulplot4)

legendpeterpaul<- get_legend(peterpaulplot4)

plotgrid1<-plot_grid(peterpaulplot2+ theme(legend.position = "none"), 
                     peterpaulplot3+ theme(legend.position = "none"), 
                     peterpaulplot4+ theme(legend.position = "none"),
                     rel_heights = c(3, 3,3), 
                     nrow = 1,
          align = "v", labels= c("Temp", "TP", "TN"), 
          label_size = 10)

plotgridfinal<-plot_grid(plotgrid1,nrow=2,legendpeterpaul, 
                         rel_heights= c(3,0.5) 
                        )
print(plotgridfinal)
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Temperature is pretty similar between both lakes over the different seasons, with mean temp reaching a low in November, and a high in Aug/Sep. Peter lake consistantly has a higher mean total phosphorus compared to Paul lake. Peter lake also has higher outliers for TP than Paul lake. The TP peaks in the summer months (7,8,9). Similarily, Peter lake has higher mean TN than paul lake. TN is very stable year around, but peaks in Aug/Sep.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
#6
Niwot_Ridge_Plot1<-
  ggplot(subset(NEON_LITTER, functionalGroup=="Needles"),
         aes(x=collectDate, y=dryMass))+
  geom_point(aes(color=nlcdClass))+
  xlab("Date")+ 
  ylab("Dry Mass")
print(Niwot_Ridge_Plot1)
#7
Niwot_Ridge_Plot2<-
  ggplot(subset(NEON_LITTER, functionalGroup=="Needles"), 
         aes(x=collectDate, y=dryMass))+
  geom_point()+
  facet_wrap(vars(nlcdClass))+
  xlab("Date")+ 
  ylab("Dry Mass")+
  theme_bw(base_size = 8)
print(Niwot_Ridge_Plot2)
```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think that the faceted plot is more effective. I think so because each class is separated out, so there is no need to color each individual data point. The lack of color helps you concentrate on comparing and contrasting different classes easily. 
