---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Taro Katayama"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
getwd()
library(tidyverse)
library(agricolae)

NTL_LTER<- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)
NTL_LTER$sampledate<-
  as.Date(NTL_LTER$sampledate, format= "%m/%d/%y")
#2
Taro_Theme <- theme_classic(base_size = 10) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "bottom")
theme_set(Taro_Theme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: there is no difference in mean lake temp with depth.
Ha: There is a difference in mean lake temp with depth. 

4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
NTL_LTERsubset<- NTL_LTER%>%
  separate(sampledate, c("Year", "Month", "Day"), sep = "-")%>%
  filter(Month == "07")%>%
  select(lakename, year4, daynum, depth, temperature_C)%>%
  na.omit()

#5
temperaturebydepth<- NTL_LTERsubset%>%
  ggplot(aes(x=depth, y=temperature_C))+
  geom_point(aes(color=lakename))+
  geom_smooth(method = lm, col="black")+
  labs(x="Depth", y="Temperature (C)")+
  ylim(0,35)+
  theme_bw()+
  theme(legend.position = "bottom")
print(temperaturebydepth)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: Generally there is a trend that temperature decreases as you increase depth. However, the residuals are quite large, which suggests that there are other variables that influence temperature. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
tempbydepthregression<-
  lm(data = NTL_LTERsubset, temperature_C~ depth)
summary(tempbydepthregression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The linear regression shows that temperature is influenced by depth. This is shown by the high R^2 value of 0.7387. It shows that around 74% of the temperature variable is explained by changes in depth. The degrees of freedom for this model is 9726. The p value is very small at <2e-16, which shows that the finding is signficant, where the null hypothesis does not hold. It is estimated that for every 1m change in depth, there is a -1.94621 change in temperature. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
NTLregression<- lm(data = NTL_LTERsubset,
                   temperature_C~depth+year4+daynum)
step(NTLregression)
#depth is best suited to predict temperature. This is because the AIC value is influenced most by depth compared to the other variables. 
#10
NTLmultipleRegression<-
  lm(data = NTL_LTERsubset, 
     temperature_C~depth+year4+daynum)

summary(NTLmultipleRegression)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggests that we use depth, year, and daynum to predict temperature. The multiple regression model shows that the R^2 value improved to a value of 0.7411. This means that the variables listed accounts for 74.11% of the change in temperature. This is a slight improvement from just using depth as the explanatory variable. The R^2 value increased from 0.7387 to 0.7411. 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
laketempsubset<-NTL_LTERsubset%>%
  group_by(lakename)%>%
  summarise(meantemp = mean(temperature_C))

laketempANOVA<- aov(data=NTL_LTERsubset, temperature_C~lakename)
summary(laketempANOVA)

laketempANOVA2<- lm(data = NTL_LTERsubset, temperature_C~lakename)
summary(laketempANOVA2)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes there is a significant difference in mean temperature among the lakes. This is shown through the p value which is <2e-16 when running the ANOVA test.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

tempbydepthLakes<-NTL_LTERsubset%>%
  ggplot(aes(x=depth, y=temperature_C))+
  geom_point(aes(color=lakename), alpha=.5, size=0.5)+
  geom_smooth(aes(group=lakename, color=lakename),
    method = lm, size=0.4, se= FALSE)+
  ylim(0,35)+
  labs(x="Depth", y="Temperature (C)")+
  theme_bw()+
  theme(legend.position = "bottom")

print(tempbydepthLakes)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(laketempANOVA)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:Statistically speaking, Paul Lake and Ward Lake have the same mean temperature as Peter Lake. 
No, although Central Long Lake is the closest, with only Crampton lake having the same mean temperature, with a p value of .066

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We can use the two sample t-test. This will compare means of just two lakes.


